**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**

### 2018-12-03 submission/manuscript.Rmd
typo
update

---
title: "**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

Here, colonic disease status is defined as Normal or Screen Relevant Neoplasias (SRN). Stool bacterial population abundances and stool hemoglobin levels of `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` Normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples were used to learn binary classifiers and evaluate their performances. The results show that ...... (Is there a maximum threshold of prediction with all these methods? Does an increase in model complexity improve predictibility?)

@@

#### Data collection
  The data used for this analysis are stool bacterial abundances, stool hemoglobin levels and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study. These data were obtained from Sze et al [@sze_leveraging_2018]. 
  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscpy. Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma. Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas. Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas. Stool provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances. The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].

@@

  To train and validate our model, labeled data is randomly split 80/20 into a training set and testing
set. Then, seven binary class classifiers, L2 logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM, decision tree, random forest and XGBoost, are learned. The training set is used for training purposes and validation of hyperparameter selection, and the test set is used for evaluation purposes. Hyperparameters are selected using 5-fold cross-validation with 100-repeats on the training set. Since the colonic disease status are not uniformly represented in the data, 5-fold splits are stratified to maintain the overall label distribution on the training set. Python programming langugae v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` functions are used to learn the seven classifiers (Table 1).

@@

**Figure 1. Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances. The boxplot shows quartiles at the box ends and the statistical median as athe horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges.

>>>>

---
title: "**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**"
bibliography: references.bib
output:
  word_document: default
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

Here, colonic disease status is defined as Normal or Screen Relevant Neoplasias (SRN). Stool bacterial population abundances and stool hemoglobin levels of `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` Normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples were used to learn binary classifiers and evaluate their performances. The results show that ...... (Is there a maximum threshold of prediction with all these methods? Does an increase in model complexity improve predictability?)

@@

#### Data collection
  The data used for this analysis are stool bacterial abundances, stool hemoglobin levels and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study. These data were obtained from Sze et al [@sze_leveraging_2018]. 
  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscopy. Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma. Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas. Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas. Stool provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances. The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].

@@

  To train and validate our model, labeled data is randomly split 80/20 into a training set and testing
set. Then, seven binary class classifiers, L2 logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM, decision tree, random forest and XGBoost, are learned. The training set is used for training purposes and validation of hyperparameter selection, and the test set is used for evaluation purposes. Hyperparameters are selected using 5-fold cross-validation with 100-repeats on the training set. Since the colonic disease status are not uniformly represented in the data, 5-fold splits are stratified to maintain the overall label distribution on the training set. Python programming language v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` functions are used to learn the seven classifiers (Table 1).

@@

**Figure 1. Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges.
